Indexing Variation Graphs\footnote{This work was supported by the Wellcome Trust grant [098051].}

### 2016-04-20 paper/paper.tex
Finished the figures; proofreading next.

% FIXME SA/BWT/LCP example
\emph{Suffix arrays} \cite{Manber1993} were introduced as a space-efficient alternative to the suffix tree. The suffix array of text $T[0, n-1]$ is an array of pointers $\SA[0, n-1]$ to the suffixes of the text in \emph{lexicographic order}. Given a text and its suffix array, we can find the occurrences of \emph{pattern} string $X$ in the text by using \emph{binary search} in $\Oh{\abs{X} \log n}$ time. The suffix array requires $n \log n$ bits of space in addition to the $n \log \sigma$ bits used by the text --- still much more than the text --- while its functionality is more limited than that of the suffix tree.

@@

We assume for convenience that all graphs have two special nodes: the \emph{source} node $s$ and the \emph{sink} node $t$. To distinguish them from the other nodes, we label them with characters $G.\glabel(s) = \#$ and $G.\glabel(t) = \$$. The label of the sink node is unique in the graph, while the label of the source node can be used in other nodes for technical purposes. We add an edge $(s, v)$ for all nodes $v \in V \setminus \set{s}$ that have no other incoming edges, and an edge $(v, t)$ for all nodes $v \in V \setminus \set{t}$ with no other outgoing edges. We also add the edge $(t, s)$ to guarantee that $G.\gindegree(v) \ge 1$ and $G.\goutdegree(v) \ge 1$ for all nodes $v \in V$. However, we assume for convenience that no path in the graph crosses the edge $(t, s)$.

@@

% FIXME example of strings, DBG, long paths, false positives
\begin{definition}[de~Bruijn graph]
Let $\mathcal{S}$ be a \kcollection{k} over alphabet $\Sigma$. The \orderk{k} \emph{de~Bruijn graph} of $\mathcal{S}$ is a graph $G = (V, E)$ such that
\begin{itemize}
\item each node $v_{X} \in V$ represent a distinct \kmer{k} $X$ occurring in $\mathcal{S}$, with $G.\glabel(v_{X}) = X[0]$;
\item each node $v_{X} \in V$ has a \emph{key} $G.\gkey(v_{X}) = X$; and
\item each edge $(v_{X}, v_{Y}) \in E$ represents a \kmer{k+1} $X[0]Y = Xc$ (where $c \in \Sigma$) occurring in $\mathcal{S}$.
\end{itemize}
We use the node corresponding to $\#^{k}$ as the source node $s$ and the node corresponding to $\$^{k}$ as the sink node $t$, adding the technical edge $(t, s)$ in the usual way.
\end{definition}

@@

% FIXME LF(i, c) picture from the slides for the earlier DBG example
The \emph{generalized compressed suffix array} (GCSA) \cite{Siren2014} is a further generalization of the XBW transform for a class of graphs that includes \emph{directed acyclic graphs} and de Bruijn graphs. Before a graph can be indexed, we have to transform it into an equivalent graph, where the nodes can be unambiguously sorted by the labels of the right-maximal paths starting from them. In the worst case, the transformation increases the size of the graph exponentially. After sorting the nodes of the transformed graph in lexicographic order, we encode them using three sequences. $\BWT$ contains the labels of the predecessors, while bitvectors $\bvIN$ and $\bvOUT$ encode the indegrees and outdegrees of each node in the same way as bitvector $B$ of the XBW transform. \LFmapping{} uses $\select$ queries on bitvector $\bvIN$ to map nodes to BWT ranges, ordinary \LFmapping{} with $\BWT$ to map incoming edges to the corresponding outgoing edges, and $\rank$ queries on bitvector $\bvOUT$ to map the outgoing edges to the predecessor nodes.

@@

As de~Bruijn graphs and \kmer{k} indexes are both based on \kmer{k}s, we can use one to simulate the other. In order to define the de~Bruijn graph of a graph, we build a \kcollection{k} based on the collection $\mathcal{S}$ of the labels of the maximal paths in graph $G = (V, E)$. If sequence $S \in \mathcal{S}$ is the label of path $P = (v_{i})_{i \in Z}$ in the graph, we set $\mathcal{S}.\gnode(S, i) = v_{i}$ for all positions $i \in Z$. We then transform $\mathcal{S}$ into a \kcollection{k} by inserting the required the number of characters $\#$ to the beginning of each non-left-infinite sequence, and characters $\$$ to the end of each non-right-infinite sequence. If $S[i]$ is a $\#$ we inserted, we set $\mathcal{S}.\gnode(S, i) = (s, j)$, where $s$ is the start node of $G$ and $j$ is the distance to the nearest non-inserted $\#$ in $S$. Similarly, if $S[i]$ is an inserted $\$$, we set $\mathcal{S}.\gnode(S, i) = (t, j)$, where $t$ is the sink node and $j$ is the distance to the nearest non-inserted $\$$.

@@

We can use a \kmer{k+1} index of graph $G = (V, E)$ to simulate the \orderk{k} de~Bruijn graph $G' = (V', E')$ of $G$. As the nodes of the de~Bruijn graph correspond to \kmer{k}s and the edges correspond to \kmer{k+1}s, we can represent the nodes by the adjacent edges. If we are interested in node $v \in V'$ with $G'.\gkey(v) = X$, we search for incoming edges $cX$ and outgoing edges $Xc$ for all $c \in \Sigma$. Determining the existence of a node requires $2 \sigma$ queries in the \kmer{k+1} index, and we learn the presence of adjacent nodes and edges in the process.

@@

We can compress a de~Bruijn graph structurally by repeatedly merging sets of nodes sharing a common prefix of their keys, as long as the conditions of Lemma~\ref{lemma:dbg-prune} hold. Let $G' = (V', E')$ be an \orderk{k} pruned de~Bruijn graph, and let $G'' = (V'', E'')$ be the same graph after further pruning. Each node $v'' \in V''$ is an \emph{equivalence class} of nodes $V'(v'') \subseteq V'$ corresponding to a shared prefix $G''.\gkey(v'')$ of keys. For all $v' \in V'(v'')$, we have $G'.\gvalue(v') = G''.\gvalue(v'')$.

@@

Path graphs are essentially equivalent to the prefix-sorted automata used in the original paper on GCSA \cite{Siren2014}. As many of the technical details are different, we now take a look at how a GCSA of a path graph works. We also extend GCSA with techniques based on suffix trees. Appendix~\ref{appendix:construction} describes a construction algorithm that builds a GCSA for a maximally pruned \orderk{2k}, \orderk{4k}, or \orderk{8k} de~Bruijn graph from a set of paths of length $k$. While the algorithm is a similar prefix-doubling algorithm as in the original GCSA, it keeps the paths and graphs on disk to avoid excessive memory usage.

@@

% FIXME refer to the figure in Section 2.4
Let $G' = (V', E')$ be a path graph. We sort the nodes $V'$ by their keys in lexicographic order and then process the nodes in that order to generate the sequences $\BWT$, $\bvIN$, and $\bvOUT$. For each node $v' \in V'$, we append
\begin{itemize}
\item $\BWT$ with the predecessor labels $G.\glabel(u')$ for all edges $(u', v') \in E'$;
\item $\bvIN$ with the indegree $x = G.\gindegree(v')$ encoded as a binary sequence $0^{x-1} 1$; and
\item $\bvOUT$ with the outdegree $x = G.\goutdegree(v')$ encoded as a binary sequence $0^{x-1} 1$.
\end{itemize}
If node $v' \in V'$ has lexicographic rank $i$, the range of incoming edges $(u', v') \in E'$ to the node is $[sp_{in}, ep_{in}] = [\bvIN.\select(i, 1) + 1, \bvIN.\select(i+1, 1)]$. The labels of predecessor nodes $u'$ are encoded in $\BWT[sp_{in}, ep_{in}]$. Sorting the incoming edges by pairs $(\BWT[j], i)$, where $\BWT[j]$ corresponds to edge $(u', v')$, is equivalent to sorting them by strings $G'.\glabel(u') \cdot G'.\gkey(v')$. As multiple edges may have the same sort keys, our sorting algorithm must be stable. We get the desired sorting order by using \LFmapping: $j \mapsto \LF(j)$.

@@

While graphs are a natural formalism for representing genetic variation, they cannot represent certain types of \emph{rearrangements} adequately. For example, if sequence $S$ can occur in different positions of the genome (e.g.~$XSYZ$ and $XYSZ$), we can have a single copy or multiple copies of $S$ in the reference. Neither of the options is good for a reference genome. With a single copy, we create paths in the reference that do not correspond to any valid genome. With the second option, we duplicate sequences that are not only identical but also functionally the same. One solution is to switch from graphs to \emph{context-free grammars}. As long as the grammar is non-nested, we can handle it with the hypertext index. We build a hypertext index for the high level graph, where each node is labeled with a nonterminal symbol, while each nonterminal expands into a subgraph indexed in GCSA.

>>>>

\emph{Suffix arrays} \cite{Manber1993} were introduced as a space-efficient alternative to the suffix tree. The suffix array of text $T[0, n-1]$ is an array of pointers $\SA[0, n-1]$ to the suffixes of the text in \emph{lexicographic order}. Given a text and its suffix array, we can find the occurrences of \emph{pattern} string $X$ in the text by using \emph{binary search} in $\Oh{\abs{X} \log n}$ time. The suffix array requires $n \log n$ bits of space in addition to the $n \log \sigma$ bits used by the text --- still much more than the text --- while its functionality is more limited than that of the suffix tree. See Figure~\ref{figure:indexes} for an example of the suffix array and related structures.

\begin{figure}[t!]
\includegraphics{gcsa2_text_indexes.pdf}
\caption{\LFmapping, BWT, suffix array, and LCP array for text $\dnaseq{GCATCATA}\$$.}\label{figure:indexes}
\end{figure}

@@

We assume for convenience that all graphs have two special nodes: the \emph{source} node $s$ and the \emph{sink} node $t$. To distinguish them from the other nodes, we label them with characters $G.\glabel(s) = \#$ and $G.\glabel(t) = \$$. The label of the sink node is unique in the graph, while the label of the source node can be used in other nodes for technical purposes. We add an edge $(s, v)$ for all nodes $v \in V \setminus \set{s}$ that have no other incoming edges, and an edge $(v, t)$ for all nodes $v \in V \setminus \set{t}$ with no other outgoing edges. We also add edge $(t, s)$ to guarantee that $G.\gindegree(v) \ge 1$ and $G.\goutdegree(v) \ge 1$ for all nodes $v \in V$. However, this edge is not considered a real edge, and no path can cross it.

@@

\begin{definition}[de~Bruijn graph]
Let $\mathcal{S}$ be a \kcollection{k} over alphabet $\Sigma$. The \orderk{k} \emph{de~Bruijn graph} of $\mathcal{S}$ is a graph $G = (V, E)$ such that
\begin{itemize}
\item each node $v_{X} \in V$ represent a distinct \kmer{k} $X$ occurring in $\mathcal{S}$, with $G.\glabel(v_{X}) = X[0]$;
\item each node $v_{X} \in V$ has a \emph{key} $G.\gkey(v_{X}) = X$; and
\item each edge $(v_{X}, v_{Y}) \in E$ represents a \kmer{k+1} $X[0]Y = Xc$ (where $c \in \Sigma$) occurring in $\mathcal{S}$.
\end{itemize}
We use the node corresponding to $\#^{k}$ as the source node $s$ and the node corresponding to $\$^{k}$ as the sink node $t$, adding the technical edge $(t, s)$ in the usual way.
\end{definition}

@@

The \emph{generalized compressed suffix array} (GCSA) \cite{Siren2014} is a further generalization of the XBW transform for a class of graphs that includes \emph{directed acyclic graphs} and de Bruijn graphs. Before a graph can be indexed, we have to transform it into an equivalent graph, where the nodes can be unambiguously sorted by the labels of the right-maximal paths starting from them. In the worst case, the transformation increases the size of the graph exponentially. After sorting the nodes of the transformed graph in lexicographic order, we encode them using three sequences. $\BWT$ contains the labels of the predecessors, while bitvectors $\bvIN$ and $\bvOUT$ encode the indegrees and outdegrees of each node in the same way as bitvector $B$ of the XBW transform. \LFmapping{} uses $\select$ queries on bitvector $\bvIN$ to map nodes to BWT ranges, ordinary \LFmapping{} with $\BWT$ to map incoming edges to the corresponding outgoing edges, and $\rank$ queries on bitvector $\bvOUT$ to map the outgoing edges to the predecessor nodes.

@@

As de~Bruijn graphs and \kmer{k} indexes are both based on \kmer{k}s, we can use one to simulate the other. In order to define the de~Bruijn graph of a graph, we build a \kcollection{k} based on the collection $\mathcal{S}$ of the labels of the maximal paths in graph $G = (V, E)$. If sequence $S \in \mathcal{S}$ is the label of path $P = (v_{i})_{i \in Z}$ in the graph, we set $\mathcal{S}.\gnode(S, i) = v_{i}$ for all positions $i \in Z$. We then transform $\mathcal{S}$ into a \kcollection{k} by inserting the required the number of characters $\#$ to the beginning of each non-left-infinite sequence, and characters $\$$ to the end of each non-right-infinite sequence. If $S[i]$ is a $\#$ we inserted, we set $\mathcal{S}.\gnode(S, i) = s:j$, where $s$ is the start node of $G$ and $j$ is the distance to the nearest non-inserted $\#$ in $S$. Similarly, if $S[i]$ is an inserted $\$$, we set $\mathcal{S}.\gnode(S, i) = (t, j)$, where $t$ is the sink node and $j$ is the distance to the nearest non-inserted $\$$.

@@

See Figure~\ref{figure:graph-dbg} for an example of a de~Bruijn graph of a graph. We can use a \kmer{k+1} index of graph $G = (V, E)$ to simulate the \orderk{k} de~Bruijn graph $G' = (V', E')$ of $G$. As the nodes of the de~Bruijn graph correspond to \kmer{k}s and the edges correspond to \kmer{k+1}s, we can represent the nodes by the adjacent edges. If we are interested in node $v \in V'$ with $G'.\gkey(v) = X$, we search for incoming edges $cX$ and outgoing edges $Xc$ for all $c \in \Sigma$. Determining the existence of a node requires $2 \sigma$ queries in the \kmer{k+1} index, and we learn the presence of adjacent nodes and edges in the process.

\begin{figure}[t!]
\includegraphics[width=396pt]{gcsa2_graph_dbg.pdf}
\caption{Left: Input graph $G = (V, E)$, with each node $v \in V$ labeled with $v:G.\glabel(v)$. Right: The \orderk{3} de~Bruijn graph $G' = (V', E')$ of graph $G$, with each node $v' \in V'$ labeled with $G'.\gkey(v')$ and $G'.\gvalue(v')$. Both: Edges $(t, s)$ are not shown. The highlighted path in the de~Bruijn graph is a false positive, as it consists of two disjoint paths in the input graph.}\label{figure:graph-dbg}
\end{figure}

@@

We can compress a de~Bruijn graph structurally by repeatedly merging sets of nodes sharing a common prefix of their keys, as long as the conditions of Lemma~\ref{lemma:dbg-prune} hold. Let $G' = (V', E')$ be an \orderk{k} pruned de~Bruijn graph, and let $G'' = (V'', E'')$ be the same graph after further pruning. Each node $v'' \in V''$ is an \emph{equivalence class} of nodes $V'(v'') \subseteq V'$ corresponding to a shared prefix $G''.\gkey(v'')$ of keys. For all $v' \in V'(v'')$, we have $G'.\gvalue(v') = G''.\gvalue(v'')$. See Figure~\ref{figure:pruned-index} for an example of a pruned de~Bruijn graph.

\begin{figure}[t!]
\includegraphics[width=396pt]{gcsa2_pruned_index.pdf}
\caption{Left: An \orderk{3} pruned de~Bruijn graph $G''$ \kequivalent{3} to the de~Bruijn graph in Figure~\ref{figure:graph-dbg}. Right: GCSA for graph $G''$. Leftward arrows show how backward searching proceeds, with the highlighted arrows demonstrating it from pattern $\dnaseq{T}$ to pattern $\dnaseq{AT}$. Rightward arrows tell where the samples for each node are located, with the highlighted ones showing them for the node with key $\dnaseq{CAT}$. The keys are not stored explicitly in the index.}\label{figure:pruned-index}
\end{figure}

@@

Path graphs are essentially equivalent to the prefix-sorted automata used in the original paper on GCSA \cite{Siren2014}. As many of the technical details are different, we now take a look at how a GCSA of a path graph works. We also extend GCSA with techniques based on suffix trees.
See Figure~\ref{figure:pruned-index} for an example of a GCSA.
Appendix~\ref{appendix:construction} describes a construction algorithm that builds a GCSA for a maximally pruned \orderk{2k}, \orderk{4k}, or \orderk{8k} de~Bruijn graph from a set of paths of length $k$. While the algorithm is a similar prefix-doubling algorithm as in the original GCSA, it keeps the paths and graphs on disk to avoid excessive memory usage.

@@

Let $G' = (V', E')$ be a path graph. We sort the nodes $V'$ by their keys in lexicographic order and then process the nodes in that order to generate the sequences $\BWT$, $\bvIN$, and $\bvOUT$. For each node $v' \in V'$, we append
\begin{itemize}
\item $\BWT$ with the predecessor labels $G.\glabel(u')$ for all edges $(u', v') \in E'$;
\item $\bvIN$ with the indegree $x = G.\gindegree(v')$ encoded as a binary sequence $0^{x-1} 1$; and
\item $\bvOUT$ with the outdegree $x = G.\goutdegree(v')$ encoded as a binary sequence $0^{x-1} 1$.
\end{itemize}
If node $v' \in V'$ has lexicographic rank $i$, the range of incoming edges $(u', v') \in E'$ to the node is $[sp_{in}, ep_{in}] = [\bvIN.\select(i, 1) + 1, \bvIN.\select(i+1, 1)]$. The labels of predecessor nodes $u'$ are encoded in $\BWT[sp_{in}, ep_{in}]$. Sorting the incoming edges by pairs $(\BWT[j], i)$, where $\BWT[j]$ corresponds to edge $(u', v')$, is equivalent to sorting them by strings $G'.\glabel(u') \cdot G'.\gkey(v')$. As multiple edges may have the same sort keys, our sorting algorithm must be stable. We get the desired sorting order by using \LFmapping: $j \mapsto \LF(j)$.

@@

While graphs are a natural formalism for representing genetic variation, they cannot represent certain types of \emph{rearrangements} adequately. For example, if sequence $S$ can occur in different positions of the genome (e.g.~$ASBC$ and $ABSC$), we can have a single copy or multiple copies of $S$ in the graph. Neither of the options is good for a reference genome. With a single copy, we create paths in the reference that do not correspond to any valid genome. With the second option, we duplicate sequences that are not only identical but also functionally the same. One solution is to switch from graphs to \emph{context-free grammars}. As long as the grammar is non-nested, we can handle it with the hypertext index. We build a hypertext index for the high level graph, where each node is labeled with a nonterminal symbol, while each nonterminal expands into a subgraph indexed in GCSA.
